Economic Theory of Fertility Explains Modern Natural Selection

### 2020-11-30 why-negative-selection.Rmd
Use shared infrastructure from import-ukbb-data; add regression controlling for risk attitude

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Overall effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
---

@@

This theory can explain why selection effects have opposite signs when AFLB is
controlled for. Suppose that people are aware of their endowments of human
capital, i.e. skills and characteristics that are valuable in the labour market.
Suppose also that education and innate human capital are complements. They
choose how long to stay in education. After leaving education they enter the
labour market and form families. If so, then people with less human capital will
leave education earlier and have their first child earlier [cf.
@caucutt2002women; @monstad2008education]. This is a pure substitution effect: the income effect plays
no role in the decision to leave education, since people are not yet earning
wages.[^no-borrow] Whilst in the labour market, higher earners will have more
children (i.e. income effects dominate). These assumptions will lead to the
pattern we observe: earnings-increasing variants will predict later first births (and
fewer children overall), but they will also predict more children within any given
age at first live birth.

@@

@@

An alternative theory is that traits selected for are linked to
externalizing behaviour and low time discount rates, via the channel of
early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
some support for this: scores which might plausibly be linked to
externalizing behaviour, like ADHD and age at smoking initiation, are
selected for. However, this theory is less good at explaining variation
in selection across the full range of scores, including physical measures, e.g.
waist-hip ratio and BMI (see also appendix Figure \@ref(fig:plot-together)). 
Externalizing behaviour also does not explain
why selection effects change sign among older parents. Overall, we believe
that the economic theory is the leading explanation.

>>>>

---
title: "Economic Theory of Fertility Explains Modern Natural Selection"
author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it. We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank. Selection effects appear consistent over time. Polygenic scores which predict lower earnings and education are selected for. Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Overall effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

@@

This theory can explain why selection effects have opposite signs when AFLB is
controlled for. Suppose that people are aware of their endowments of human
capital, i.e. skills and characteristics that are valuable in the labour market.
Suppose also that education and innate human capital are complements. Individuals
choose how long to stay in education; after leaving education they enter the
labour market and form families. If so, then people with less human capital will
leave education earlier and have their first child earlier [cf.
@caucutt2002women; @monstad2008education]. This is a pure substitution effect: the income effect plays
no role in the decision to leave education, since people are not yet earning
wages.[^no-borrow] Whilst in the labour market, higher earners will have more
children (i.e. income effects dominate). These assumptions will lead to the
pattern we observe: earnings-increasing variants will predict later first births (and
fewer children overall), but they will also predict more children within any given
age at first live birth.

@@

```{r calc-risk-attitude}

drake::loadd(res_risk_control)
drake::loadd(res_all)

n_risk_sig <- res_risk_control %>% 
            filter(term == 'f.2040.0.0') %>% 
            {sum(.$p.value <= 0.05/33)}

n_pgs_sig <- res_risk_control %>% 
            filter(term == score_name) %>% 
            {sum(.$p.value <= 0.05/33)}

res_compare <- res_all %>% 
  filter(dep.var=='N children', reg.type == 'raw') %>% 
  left_join(res_risk_control %>% filter(term == score_name), by = 'score_name')

ratio <- res_compare$estimate.y/res_compare$estimate.x
n_less_sig <- sum(res_compare$p.value.y <= 0.05/33 & res_compare$p.value.x > 0.05/33)
```

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and age at smoking initiation, are selected for. However, this theory
is less good at explaining variation in selection across the full range of
scores, including physical measures, e.g. waist-hip ratio and BMI (see also
appendix Figure \@ref(fig:plot-together)). Externalizing behaviour also does not
explain why selection effects change sign among older parents.

We test the alternative theory directly by re-estimating equation
\@ref(eq:regression) controlling for a measure of risk attitude (Biobank field
2040). While risk attitude is always a highly significant predictor of number of
children, it has little impact on the effect size or significance of the PGS.
The median ratio of PGS effect sizes between regressions with and without
controls is `r median(ratio)`; all PGS which are significant at $p \le 0.05/33$
in uncontrolled regressions remain so when controlling for risk attitude. Thus,
although risk attitude does predict number of children, it appears not to
mediate selection effects. Overall, we believe that the economic theory is the
leading explanation.

@@
